---
title: "Data Visualization: `ggplot` part 3"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Adjust the aesthetics of plots in ggplot using `themes`, `RColorBrewer`, and `paletteer.`  
2. Use faceting to produce multi-panel plots.  

## Group project  
We will meet in our groups today to compare the data sets that we found. You should try to pick the data set that you think will be most appropriate; i.e. does it have a mix of variables, does it have a story to tell? Once you have a data set, think about the kinds of questions you might ask. Remember, in the end you will need to deliver a 5-6 minute presentation of your work. If you need help, please ask!  

## Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [R for Data Science](https://r4ds.had.co.nz/)
- [R Cookbook](http://www.cookbook-r.com/)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)
- [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3)

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(RColorBrewer)
library(paletteer)
library(janitor)
library(here)
```

```{r}
options(scipen=999) #cancels the use of scientific notation for the session
```

## Data
For this tutorial, we will use:  
[Homerange](http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1). The data are from: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
```{r message=FALSE, warning=FALSE}
homerange <- read_csv(here("lab11", "data", "Tamburelloetal_HomeRangeDatabase.csv"))
```

## ggplot themes
There are many option to change the theme of your plots. Have a look [here](https://www.datanovia.com/en/blog/ggplot-themes-gallery/) for some excellent examples. I am going to build a basic plot here and then use it throughout to illustrate the themes.
```{r}
p <- homerange %>% 
  ggplot(aes(x=taxon, fill=trophic.guild))+
  geom_bar(na.rm=T, position="dodge")
p
#Plotting counts of different taxa
```

Now if you want to test a theme, all you have to do is add it to p!

```{r}
p+theme_linedraw()+
  theme(axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Legends
There are lots of options to manipulate legends. Have a look [here](http://www.sthda.com/english/wiki/ggplot2-legend-easy-steps-to-change-the-position-and-the-appearance-of-a-graph-legend-in-r-software).
```{r}
p+theme_linedraw()+
  theme(legend.position = "bottom", #By defualt, legend will be "right"
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```
```{r}
p+theme_linedraw()+
  theme(legend.position = 'none', #By defualt, legend will be "right"
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
#'none' gets rid of legend
```

## Not enough? [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/)
```{r}
#install.packages("ggthemes")
library(ggthemes)
```

Here is a list of the `ggthemes`
```{r}
ls("package:ggthemes")[grepl("theme_", ls("package:ggthemes"))]
```

```{r}
p+ theme_pander()+ #The themes don't always work together with the adjustments well
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## RColorBrewer
The default colors used by ggplot are often uninspiring. They don't make plots pop out in presentations or publications, and you may want to use a customized palette to make things visually consistent.  

Access the help for `RColorBrewer`.
```{r}
?RColorBrewer
#3 palettes: seqential, diverging, and qualitative. Those palettes are meant for those kinds of data
```

The thing to notice is that there are three different palettes: 1) sequential, 2) diverging, and 3) qualitative. Within each of these there are several selections. You can bring up the colors by using `display.brewer.pal()`. Specify the number of colors that you want and the palette name.
```{r}
display.brewer.pal(8,"RdGy")
#Pick a palette, type in number of variables you intend on including (8 color palette for PRGn)
```

The [R Color Brewer](http://colorbrewer2.org/#type=sequential&scheme=BuGn&n=3) website is very helpful for getting an idea of the color palettes. To make things easy, use these two guidelines:

+`scale_colour_brewer()` is for points  (geom_point)
+`scale_fill_brewer()` is for fills  (geom_bar)

Here I chose the `Paired` palette. Take a moment and experiment with other options.
```{r}
p+scale_fill_brewer(palette = "Paired")+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

Once you have p+scale_fill_brewer(palette = "Paired")+, use this for the rest of your graphs to make them look consistent!


## Manually Setting Colors
You can also use `paleteer` to build a custom palette for consistency. To access the `paleteer` collection, I add it to a new object.
```{r}
colors <- paletteer::palettes_d_names
view(colors)
```

Now we can display the palettes. Assign the palette to `my_palette` and then build this base R bar plot. There are a lot of options; `paleteer` is a collection of popular palettes. I really like the [`ggsci` package] (https://cran.r-project.org/web/packages/ggsci/vignettes/ggsci.html)
```{r}
my_palette <- paletteer_d("ggsci::hallmarks_light_cosmic")
```

```{r}
barplot(rep(1,14), axes=FALSE, col=my_palette)
#Looking at colors from pallete in this base r command!
```

Now we just identify `my_palette` as part of `scale_fill_manual()`
```{r}
p+scale_fill_manual(values=my_palette)+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 60, hjust=1))+
  labs(title = "Observations by Taxon in Homerange Data",
       x = NULL,
       y= "n",
       fill= "Trophic Guild")
```

## Adjusting the x and y limits
There are many options for adjusting the x and y axes. For details, look over examples in the [R Cookbook](http://www.cookbook-r.com/). To adjust limits, we can use the `xlim` and `ylim` commands. When you do this, any data outside the specified ranges are not plotted.
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = locomotion)) +
  geom_point()
#What if we're only interested in a certian area of the plot or point? xlim and ylim
```

```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = locomotion)) +
  geom_point() +
  xlim(0, 4) +
  ylim(1, 6)
#Warning is there because there's that many values that aren't in your lim plot
```
```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = locomotion)) +
  geom_point() +
  annotate("text", label="r^2==0", parse=T, x=6, y=0)
#Parse converts special characters. Need to tell it where you want to put it by identifying X and Y axis
```

## Practice
1. Revisit the relationship between mass and homerange using a scatterplot, but color the points by taxonomic class using a custom color theme.
```{r}
glimpse(homerange)
```

```{r}
homerange %>% 
  ggplot(aes(x = log10.mass, y = log10.hra, color = class))+
  geom_point()+
  scale_color_brewer(palette = "BrBG")
```

2. Build bar plots that show the representation by taxonomic class with carnivores and herbivores. Try using a color theme that matches your scatterplot to keep everything consistent.
```{r}
homerange %>% 
  ggplot(aes(x = trophic.guild, y = class, fill = class))+
  geom_col(position="dodge", alpha = 0.75, color = "black")+
  scale_fill_brewer(palette = "BrBG")
```

## Faceting: `facet_wrap()`
[Faceting](https://ggplot2-book.org/facet.html) is one of the amazing features of ggplot. It allows us to make multi-panel plots for easy comparison. Here, I am making histograms of log10.mass for every taxon. We read the `~` in the `facet_wrap()` layer as `by`. The bin number doesn't work for every taxon, but it is enough to get the idea.
```{r}
p2 <- homerange %>% 
  ggplot(aes(x = log10.mass)) +
  geom_density()
#Basically building a series of plots stacked all at once
#So we're doing density distribution for now
#How would you make desnity plots for every animal? You'd have to filter for all of them... BUT thanks to facet, not really!
```

```{r}
p2+ facet_wrap(~taxon)
#~ =  "by", so "wrap by taxon"
#You get density plots for all the different data at once
```
```{r}
p2+ facet_wrap(~class)
```


Don't give R a categorical variable that has hundreds of different values! Ex. species. TONS of things to generate!


## Faceting: `facet_grid()`
As you can imagine, there are lots of options for faceting. Another useful facet type is `facet_grid`. 
```{r}
p2+facet_grid(thermoregulation~class)
#So two caterogical variables!
#Use facet_grid for 2 categorical variables
```


##Printing
There are many options to export plots for use outside of R. Here is the basic code to do exports.
```{r}
pdf("homerange.pdf") #pdf(file, then file type)
p2 #Name of plot you want to export (stored as an object)
dev.off() #Then close the device (dev.off = close it)
#Need to tell R what you want to export
```



## That's it, let's take a break!   

-->[Home](https://jmledford3115.github.io/datascibiol/)