---
title: "Transforming data 2: `filter()`"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
    keep_md: yes
  pdf_document:
    toc: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Use the functions of dplyr (filter, select, arrange) to organize and sort data frames.  
2. Use `mutate()` to calculate a new column from existing columns.  

## Review  
In the previous lab, we used `select()` to extract columns of interest from a data frame. This helps us focus our attention on the variables relevant to our question. However, it doesn't allow us to extract observations from within the data frame. The `filter()` function allows us to extract data that meet specific criteria. When combined with `select()`, we have the power to transform, shape, and explore data with the potential to make new discoveries. 

#Filter works with rows
Power to relimit data based off of columns and rows. 

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library("tidyverse")
```

## Load the data
For these exercises we will use the data from lab 4_1.  
```{r}
fish <- readr::read_csv("data/Gaeta_etal_CLC_data.csv")
mammals <- readr::read_csv("data/mammal_lifehistories_v2.csv")
```

To keep things tidy, I am going to rename the variables in the mammals data.  
```{r}
mammals <- rename(mammals, genus=Genus, wean_mass="wean mass", max_life="max. life", litter_size="litter size", litters_per_year="litters/year")
```

## `filter()`
Unlike `select()`, `filter()` allows us to extract data that meet specific criteria within a variable. Let's say that we are interested only in the fish that occur in lake "AL". We can use `filter()` to extract these observations.  
```{r}
filter(fish, lakeid == "AL")
#Fish is dataframe. Lakeid is restricted to AL
#One of the variable frames is AL, pull out dataframes for AL, or Alabama. Down to 383 obersations. Filtered out observations where lakeid is limited to AL
#If you still had columns renamed from your previous environment, then error because you already renamed. So instead erase global environment and rename/restart again
```

Similarly, if we are only interested in fish with a length greater than or equal to 350 we can use `filter()` to extract these observations.  
(Using these as operators. Searching for all observations in length column that are greater than or equal to 350)
(Select is looking within columns. Filter is looking within rows)
```{r}
filter(fish, length >= 350)
```

+ `filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).  

Using the `!` operator allows for the exclusion of specific observations.
(! is the not operator. So everything that is not AL operator)
```{r}
filter(fish, lakeid != "AL")
```

## Using `filter()` with multiple observations  
Filtering multiple values within the same variable requires the `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm).    
(Filtering fish. Inside length column, 167 and 175. Length is either 167 or 175. So filtering out 2 potential targets)
```{r}
filter(fish, length %in% c(167, 175))
```

Alternatively, you can use `between` if you are looking for a range of specific values.  
(All of the observations are either between 2.5 or 2.55)
```{r}
filter(fish, between(scalelength, 2.5, 2.55))
```

You can also extract observations "near" a certain value but you need to specify a tolerance.  
(Filter fish. Want to filter radii length in millimeters, and filter values near 2, but tolerance is 0.2)
```{r}
filter(fish, near(radii_length_mm, 2, tol = 0.2))
```

## Practice
1. Filter the `fish` data to include the samples from lake "BO".
```{r}
filter(fish, lakeid == "BO")
```

2. Filter the data to include all lakes except "AR".
```{r}
filter(fish, lakeid != "AR")
```

3. Filter the fish data to include all fish with a scalelength within 0.25 of 8.
```{r}
filter(fish, near(scalelength, 8, tol = 0.25))
```

4. Filter the fish data to include fish with a scalelength between 2 and 4.

```{r}
filter(fish, between(scalelength, 2, 4))
```
(If something works, you get very excited. BUT just because it works doesn't mean that it's correct. Run various checks. Ex. To check for this)
```{r}
x <- filter(fish, between(scalelength, 2, 4))
View(x)
```
(From here on out, we're using filter and select exclusively)

## Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract only the fish that have lakeid "AL" and length >350.
```{r}
filter(fish, lakeid == "AL" & length > 350)
#Lake ID equal to AL and length is greater tan 350
#When you put & in there, it looks through rows where lake ID is AL AND length is 350, so both conditions must be met for this to be satisfactory
```
```{r}
filter(fish, lakeid == "AL" & length == 350)
```

Notice that the `|` operator generates a different result.
```{r}
filter(fish, lakeid == "AL" | length == "350")
#Meant to be an "or" value. 
#Do we predict that in this result, we will see lake IDs other than AL? Yes, because looking for EITHER of those conditions. Either length is 350 or lakeid is AL
```
```{r}
y <- filter(fish, lakeid == "AL" | length > "350")
view(y)
```

Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met.  
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 and/or condition 2 is met.  
+ `filter(xor(condition1, condition2)` will return all rows where only one of the conditions is met, and not when both conditions are met.  

In this case, we filter out the fish with a length over 400 and a scale length over 11 or a radii length over 8.
```{r}
filter(fish, length > 400, (scalelength > 11 | radii_length_mm > 8))
```

## Practice  
1. Have a look at the mammals data using the summary functions of your choosing.    
```{r}
glimpse(msleep)
```

2. What are the names of the variables in the mammals data?  
```{r}
names(msleep)
```

3.  From the `mammals` data, filter all members of the family Bovidae with a mass greater than 450000.
```{r}
filter(mammals, family == "Bovidae" & mass >= "450000")
```
```{r}
#Professor's way
bovidae <- filter(mammals, family == "Bovidae" & mass >= "450000")
```

4. Let's say we are only interested in family, genus, species, mass, and gestation. Relimit the mammals data frame to these variables.  
```{r}
select(mammals, "family", "genus", "species", "mass", "gestation")
```
```{r}
#Professor's way:
mammals_relimited <- select(mammals, "family", "genus", "species", "mass", "gestation")
mammals_relimited
```

5. Use the output from #4 to focus on the family Bovidae with a mass greater than 450000.
```{r}
bovidae <- filter(mammals_relimited, family == "Bovidae" & mass >= "450000")
bovidae
```

6. From the `mammals` data, build a data frame that compares `mass`, `gestation`, and `newborn` among the primate genera `Lophocebus`, `Erythrocebus`, and `Macaca`. Among these genera, which species has the smallest `newborn` mass?
```{r}
primate_genera <- filter(mammals, genus %in% c("Lophocebus","Erythrocebus","Macaca"))
primate_genera
```

```{r}

```

## Wrap-up  

Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)