---
title: "Lab 7 Homework"
author: "Please Add Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(skimr)
``` 

## Data
**1. For this homework, we will use two different data sets. Please load `amniota` and `amphibio`.**  

`amniota` data:  
Myhrvold N, Baldridge E, Chan B, Sivam D, Freeman DL, Ernest SKM (2015). “An amniote life-history
database to perform comparative analyses with birds, mammals, and reptiles.” _Ecology_, *96*, 3109.
doi: 10.1890/15-0846.1 (URL: https://doi.org/10.1890/15-0846.1).
```{r}
amniota <- readr::read_csv("data/amniota.csv")
```

`amphibio` data:  
Oliveira BF, São-Pedro VA, Santos-Barrera G, Penone C, Costa GC (2017). “AmphiBIO, a global database
for amphibian ecological traits.” _Scientific Data_, *4*, 170123. doi: 10.1038/sdata.2017.123 (URL:
https://doi.org/10.1038/sdata.2017.123).
```{r}
amphibio<- readr::read_csv("data/amphibio.csv")
```

## Questions  
**2. Do some exploratory analysis of the `amniota` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  

```{r}
summary(amniota)
```

The negative -999 in the data indicate that there are NAs

```{r}
amniota %>% 
  skimr::skim()
```
```{r}
#Curious where -30259 came from, maybe multiple -999 recordings?
-30259/-999
```

The NAs in this dataset seem to be represented by -999.


```{r}
amniota_no_nas <- amniota %>%
  na_if("-999")
summary(amniota_no_nas)
```


```{r}
#Trying to figure out what -30258.71100 is
amniota_no_nas %>%
  filter(female_maturity_d >= 50) %>%
  tabyl(female_maturity_d)
#this might be what you were trying to do?
```
```{r}
#What Min-Yao suggested for getting rid of that one negative value:
amniota_na <- amniota %>% 
  na_if("-999") %>% #converts all -999 values to NA
  mutate(female_maturity_d_new = ifelse(female_maturity_d < 0, NA, female_maturity_d)) %>% 
  select(-female_maturity_d)

amniota_na %>% 
  summarize(number_nas = sum(is.na(amniota_na)))
```



**3. Do some exploratory analysis of the `amphibio` data set. Use the function(s) of your choice. Try to get an idea of how NA's are represented in the data.**  

```{r}
summary(amphibio)
```
```{r}
glimpse(amphibio)
```

The data for the amphibio dataset seems to be represented by NAs.

```{r}
hist(amphibio$Aqu)
```

Uh oh...

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

```{r}
naniar::miss_var_summary(amphibio)
```


**4. How many total NA's are in each data set? Do these values make sense? Are NA's represented by values?**   

```{r}
amniota %>% 
  summarize(number_nas = sum(is.na(amniota)))
```
No, the values don't make sense for amniota.
```{r}
amphibio %>% 
  summarize(number_nas = sum(is.na(amphibio)))
```
Yes, in the case of amphibio. (Or it might be a bit big for the dataset? There were 0s in that and someone on Discord suggested the 0s were included as literally "not applicable" values as well)

**5. Make any necessary replacements in the data such that all NA's appear as "NA".**   
```{r}
#What Min-Yao suggested for getting rid of that one negative value:
amniota_na <- amniota %>% 
  na_if("-999") %>% #converts all -999 values to NA
  mutate(female_maturity_d_new = ifelse(female_maturity_d < 0, NA, female_maturity_d)) %>% 
  select(-female_maturity_d)

amniota_na %>% 
  summarize(number_nas = sum(is.na(amniota_na)))
```

```{r}
amphibio_na <- amphibio %>%
  na_if("0")
amphibio_na %>% 
  summarize(number_nas = sum(is.na(amphibio_na)))
```

**6. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amniota` data.**  
```{r}
naniar::miss_var_summary(amniota_na)
```

**7. Use the package `naniar` to produce a summary, including percentages, of missing data in each column for the `amphibio` data.**
```{r}
naniar::miss_var_summary(amphibio_na)
```

**8. For the `amniota` data, calculate the number of NAs in the `egg_mass_g` column sorted by taxonomic class; i.e. how many NA's are present in the `egg_mass_g` column in birds, mammals, and reptiles? Does this results make sense biologically? How do these results affect your interpretation of NA's?**  

```{r}
amniota_na %>% 
  select(class, egg_mass_g) %>%
  naniar::miss_var_summary(order=T)
```


**9. The `amphibio` data have variables that classify species as fossorial (burrowing), terrestrial, aquatic, or arboreal. Calculate the number of NA's in each of these variables. Do you think that the authors intend us to think that there are NA's in these columns or could they represent something else? Explain.**
```{r}
amphibio
```
```{r}
amphibio_na %>% 
  select(Fos, Ter, Aqu, Arb) %>%
  naniar::miss_var_summary(order=T)
```

I think that the NAs could also mean exactly that: 'Not applicable." Some species aren't aquatic, others aren't terrestrial and so on, so they don't apply to certain classifications.

**10. Now that we know how NA's are represented in the `amniota` data, how would you load the data such that the values which represent NA's are automatically converted?**
```{r}
amniota_converted_na <- 
  readr::read_csv(file = "data/amniota.csv", na = c("-999"))
```
</div>

## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.  