---
title: "Midterm review and `naniar()`"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float: yes
    keep_md: yes
  pdf_document:
    toc: yes
---

## Learning Goals
*At the end of this exercise, you will be able to:*    
1. Identify and manage NA's is data sets where NA's are represented in multiple ways.    
2. Use the `naniar` and `visdat` packages to help manage NA's in large data sets.  

## Breakout Rooms  
Please take 5-8 minutes to check over your answers to HW 6 in your group. If you are stuck, please remember that you can check the key in [Joel's repository](https://github.com/jmledford3115/BIS15LW2021_jledford).  

## Midterm 1 Review
Let's briefly review the questions from midterm 1 so you can get an idea of how I was thinking about the problems. Remember, there is more than one way to get at these answers, so don't worry if yours looks different than mine!

(Code chunks are going to start to be getting bigger, so more room for mistakes, so please be sure to reach out for help if you need it!)

## Load the libraries
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
library("skimr")
```

## Review
When working with outside or "wild" data, dealing with NA's is a fundamental part of the data cleaning or tidying process. Data scientists spend most of their time cleaning and transforming data- including managing NA's. There isn't a single approach that will always work so you need to be careful about using replacements strategies across an entire data set. And, as the data sets become larger NA's can become trickier to deal with.  

For the following, we will use life history data for mammals. The [data](http://esapubs.org/archive/ecol/E084/093/) are from:  
**S. K. Morgan Ernest. 2003. Life history characteristics of placental non-volant mammals. Ecology 84:3402.**  

## Practice
1. Load the mammals life history data and clean the names.  
```{r}
life_history <- readr::read_csv("data/mammal_lifehistories_v3.csv")
life_history <- janitor::clean_names(life_history) #Cleaned names
```

2. Use one or more of the functions from the last lab to determine if there are NA's in the data, how they are represented, and where they are located.
```{r}
life_history %>% 
  skimr::skim()
#glimpse works as well, same thing as summary. "Not recorded" could count as NAs as well.
```

```{r}
life_history_no_nas <- life_history %>%
  na_if("-999")
```

```{r}
life_history_no_nas %>% 
  summarize(number_nas=sum(is.na(life_history_no_nas)))
```
```{r}
life_history_no_nas %>% 
  purrr::map_df(~ sum(is.na(.)))
#Want to map to a dataframe, sum is.na to all cells
```
```{r}
hist(life_history_no_nas$wean_mass)
```
```{r}
life_history_no_nas %>% 
  purrr::map_df(~ sum(is.na(.))) %>% #Want to map to a dataframe, sum is.na to all cells
  pivot_longer(everything(),
    names_to= "variables",
    values_to = "num_nas") %>% 
  arrange(desc(num_nas))
```

(One way NAs are represented: -999.
Another way is as NAs as well.)


3. Can we use a single approach to deal with NA's in this data set? Given what you learned in the previous lab, how would you manage the NA values?

(Na.if command is helpful, but doesn't deal with everything, so need to handle things in a case-by-case basis)
```{r}
life_history_no_nas %>% 
  na_if("-999") %>% 
  mutate(newborn=na_if(newborn, ("not measured"))) %>% #Replacing in newborn, not measured with NA
  purrr::map_df(~sum(is.na(.)))
#This all shows the NAs throughout the data
```


## `naniar`
`naniar` is a package that is built to manage NA's. Many of the functions it performs can also be performed using tidyverse functions, but it does provide some interesting alternatives.

`miss_var_summary` provides a clean summary of NA's across the data frame.
```{r}
naniar::miss_var_summary(life_history_no_nas)
#Always look at the output of the data and ask yourself: "Does this data make sense?"
#Ex. Pg. 2, max life is measurement of maximum lifespan. Is it reasonable that we know the lifespans of all of the mammals in the data? Probably not. Might require a little bit of exploration.
#HAVE to think like a biologist when looking at data!
```

Notice that `max_life` has no NA's. Does that make sense in the context of this data?
```{r}
hist(life_history_no_nas$max_life)
#Finding examples where NAs are represented by values that they shouldn't be
```

```{r}
life_history_no_nas <- 
  life_history_no_nas %>% 
  mutate(max_life=na_if(max_life, 0))
```

```{r}
naniar::miss_var_summary(life_history_no_nas)
#Now max life is fixed. Columns represent individuals they looked at
```

We can also use `miss_var_summary` with `group_by()`. This helps us better evaluate where NA's are in the data.
```{r}
life_history_no_nas %>%
  group_by(order) %>%
  select(order, wean_mass) %>%  #Selecting two variables order and wean_mass
  naniar::miss_var_summary(order=T)
#Seeing percentage of missing data. Ex. for whales, it makes sense we don't know the mass of a baby that stopped feeding from its mother, so biologically it makes sense
```

`naniar` also has a nice replace function which will allow you to precisely control which values you want replaced with NA's in each variable.
```{r}
life_history %>% 
  naniar::replace_with_na(replace = list(newborn = "not measured", weaning= -999, wean_mass= -999, afr= -999, max_life= 0, litter_size= -999, gestation= -999, mass= -999)) %>%
  naniar::miss_var_summary()
#Create a list of all the variables and show what you want them to be replaced with. Same output as before, but may or may not be clearer than before depending on how many NAs are treated in the data
```

## Practice
Let's practice evaluating NA's in a large data set. The data are compiled from [CITES](https://cites.org/eng). This is the international organization that tracks trade in endangered wildlife. You can find information about the data [here](https://www.kaggle.com/cites/cites-wildlife-trade-database).  

Some key information:  
[country codes](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2)  

1. Import the data and do a little exploration. Be sure to clean the names if necessary.
```{r}
cites <- readr::read_csv("data/cites.csv")
cites <- janitor::clean_names(cites)
cites$year <- as.factor(cites$year)
```

```{r}
cites %>% 
  skimr::skim()
```
```{r}
summary(cites)
```



2. Use `naniar` to summarize the NA's in each variable.
```{r}
naniar::miss_var_summary(cites)
```

3. Try using `group_by()` with `naniar`. Look specifically at class and `exporter_reported_quantity`. For which taxonomic classes do we have a high proportion of missing export data?
```{r}
cites %>%
  group_by(class) %>%
  select(class, exporter_reported_quantity) %>%
  naniar::miss_var_summary(order=T) %>% 
  arrange(desc(pct_miss))
#Implies that gov isn't keeping complete track of what they're exporting
#Ex. not keeping track of endangered species being exported. Could be a point to bring up in a report!
```



## Visualizing NAs
There is another package `visdat` that can be used to visualize the proportion of different classes of data, including missing data. But, it is limited by size.
```{r}
library(visdat)
```

```{r}
vis_dat(life_history_no_nas) #classes of data
```

```{r}
vis_miss(life_history_no_nas)
#Proportion of straight up missing data
```

## Dealing with NA's in advance
If you are sure that you know how NA's are treated in the data, then you can deal with them in advance using `na()` as part of the `readr` package.

(If you use this approach, you'd BETTER know how NAs are reported in your value! If you do this at the very beginning, it can lead to a lot of errors downstream. BUT using na = c enables you to deal with all of the NAs at once. BUT sometimes you'll still have issues, so sometimes you'll have to deal with variables case by case.)
```{r}
life_history_advance <- 
  readr::read_csv(file = "data/mammal_lifehistories_v3.csv", 
                  na = c("NA", " ", ".", "-999")) #all NA, blank spaces, .,and -999 are treated as NA
```

Note you would still need to deal with some variables case-by-case.
```{r}
naniar::miss_var_summary(life_history_advance)
```

(This is a very important part of the data cleaning process!)


## Wrap-up  
Please review the learning goals and be sure to use the code here as a reference when completing the homework.  
-->[Home](https://jmledford3115.github.io/datascibiol/)